---
title: "Using distributional semantics to annotate for semantic predictors"
subtitle: "A case study on the English dative alternation"
author:
  - Chiara Paolini
  - Benedikt Szmrecsanyi
  - Mariana Montes
format:
  revealjs:
    footer: "BKL - Liège, 2022-10-21"
execute:
  echo: false
---

```{r}
#| label: setup
#| include: false
headings <- c("Introduction: the dative alternation",
              "From vectors to predictors",
              "Forests and clouds",
              "Insights and further research")
source(here::here("BKL_22", "plots.R"))
```

## Outline

```{r}
#| label: outline
#| results: asis
for (heading in headings) {
  cat("-", heading, "\n")
}
i <- 1
```

# `r headings[[i]]`

## The dative alternation

::: {.footer}
`r headings[[i]]`
:::

## Modelling grammatical alternations

Traditional vs distributional predictors.

::: {.footer}
`r headings[[i]]`
:::

```{r}
#| label: add-heading1
#| include: false
i <- i+1
```

# `r headings[[i]]`

## Automatically-generated semantic predictors

Define distributional semantics.

::: {.footer}
`r headings[[i]]`
:::

## Matrix of recipients

::: {.footer}
`r headings[[i]]`
:::

## Clustering

::: {.footer}
`r headings[[i]]`
:::


```{r}
#| label: add-heading2
#| include: false
i <- i+1
```

# `r headings[[i]]`

## Cloud of recipients

```{r}
#| label: fig-tsne-r
#| fig-cap: Scatterplot of recipient lemmas. Colors are clusters, labelled by their medoids.
tsne_plot(recipient_coords) %>%
  add_point(recipient_coords %>% filter(cw == "government/nn"), -6, 0) %>%
  add_point(recipient_coords %>% filter(cw == "mother/nn"), -7)
```

::: {.footer}
`r headings[[i]]`
:::

## Cloud of themes

```{r}
#| label: fig-tsne-t
#| fig-cap: Scatterplot of themes lemmas. Colors are clusters, labelled by their medoids.
tsne_plot(theme_coords) %>%
  add_point(theme_coords %>% filter(cw == "it/pp"), -7, -10, vjust = 1)
```

::: {.footer}
`r headings[[i]]`
:::

## Random forest of distributional predictors

```{r}
#| label: fig-rf-dist
#| fig-cap: Variable importances from a random forest with only distributional predictors.
varimp_dotplot(varimp_bin)
```

::: {.footer}
`r headings[[i]]`
:::

## Random forest of traditional and distributional predictors

Still need to create the forest. 

::: {.footer}
`r headings[[i]]`
:::
